Notes on Hartree-Fock Stability of HEG

### 2016-07-01 overview/heg_equations.tex

\section{Equations}

\subsection{Homogeneous Electron Gas}

@@

\subsection{Hartree-Fock Stability}
The HF stability conditions (to my knowledge) were presented by Thouless \cite{Thouless1972} ($1^{st}$ Ed. 1960). More general conditions were presented by Adams\cite{Adams1962}. The stability condition is that the eigenvalues ($\lambda$) given by (p. 41 of \cite{Thouless1972})

@@

\begin{equation}
(\epsilon_a-\epsilon_i)C_{ai}^* + \sum\limits_{\text{j}}^{occ}\sum\limits_{\text{b}}^{vir}
	(\braket{ij|ab}-\braket{ij|ba})C_{bj}
	+(\braket{ib|aj}-\braket{ib|ja})C_{bj}^*
	=\lambda C_{ai}^*,
\end{equation}
are nonnegative. This is equivalent to the RPA oscillation frequencies being all real (p. 115 of \cite{Thouless1972}). The RPA is a linearized approximation to Time-Dependent Hartree-Fock theory. These can be written\cite{Dunning1967}\cite{Seeger1977} in the following form:
\begin{equation}
{}^{1,3}\bf{H'}=
\begin{bmatrix}
{}^{1,3}\bf{A'} & {}^{1,3}\bf{B'} \\
\left({}^{1,3}\bf{B'}\right)^* & \left({}^{1,3}\bf{A'}\right)^* \\
\end{bmatrix}
\end{equation}
Where the $1,3$ denote singlet and triplet states, respectively. $\bf{A}$, $\bf{B}$ have dimension $N_{occ}\times N_{vir}$ and are defined as follows:
\begin{subequations}
	\begin{align}
	{}^{1}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} + 2\braket{aj|ib}-\braket{aj|bi}\\
	{}^{3}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} - \braket{aj|bi}\\
	{}^{1}{B'}_{i\rightarrow a, j\rightarrow b} &= 2\braket{ab|ij}-\braket{ab|ji}\\
	{}^{3}{B'}_{i\rightarrow a, j\rightarrow b} &= -\braket{ab|ji}
	\end{align}
\end{subequations}
\begin{figure}[h]
   \includegraphics[width=1.0\linewidth]{figures/seeger_stability.pdf}
  	\caption {Hartree-Fock Stability conditions reproduced from Seeger \& Pople \cite{Seeger1977}}
  	\label{fig:stability_table}
\end{figure}

\subsection{Davidson's Algorithm}
\subsubsection{Theory}
Oftentimes we're interested only in the lowest or highest few eigenvalues of a large, symmetric matrix. Matrix eigenvalue problem operation counts scale as \url{~}$O(N^3)$ for an $N\times N$ matrix, while the memory requirement scales quadratically. Iterative subspace methods based off the Rayleigh-Ritz method are popular ways of dealing with this issue. The most famous are the Lanczos (hermitian) and Arnoldi (non-hermitian) algorithms, which build as their subspace a Krylov space. A more involved algorithm was proposed by Davidson\cite{Davidson1975} and involves building a subspace with correction vectors to be described momentarily. But first, a few definitions are introduced to define the notation to be used.

@@

\subsubsection{Implementation}
\begin{algorithm}[H]
\caption{Davidson's Algorithm}\label{davidson}
\begin{algorithmic}[1]
\State create \bf{V}
\For {$i = 0, max\text{ }iterations$}
\State $\bf{\tilde{A}} = \bf{V}^\dag\bf{AV}$
\State $\tilde{\lambda},\mathbf{\tilde{x}} = eigensystem(\bf{\tilde{A}})$
\State sort $\tilde{\lambda},\mathbf{\tilde{x}}$
\If {$i > \textit{stringlen}$} \Return false
\EndIf
\State $j \gets \textit{patlen}$
\If {$\textit{string}(i) = \textit{path}(j)$}
\State $j \gets j-1$.
\State $i \gets i-1$.
\State \textbf{goto} \emph{loop}.
\State \textbf{close};
\EndIf
\State $i \gets i+\max(\textit{delta}_1(\textit{string}(i)),\textit{delta}_2(j))$.
\State \textbf{goto} \emph{top}.
\EndFor

>>>>

\section{Homogeneous Electron Gas}

@@

The analytic form of the exchange energy is known in the continuous case. Let $y = \frac{k}{k_{F\sigma}}$. Equations \ref{eq:hfexch} through \ref{eq:fnd} are from p. 81-82 of \cite{Guiliani2005}. In 2 or 3 dimensions the exchange energy is,

\begin{equation}\label{eq:hfexch}
\epsilon_{k\sigma}^{(x)} = -\frac{2e^2k_{F\sigma}}{\pi}f_{ND}\left(\frac{k}{k_{F\sigma}}\right),
\end{equation}
where in 3D, the function $f_{3D}$ is given by
\begin{equation}
f_{3D}\left(y\right) = \frac{1}{2} + \frac{1-y^2}{4y} ln \left|\frac{1+y}{1-y} \right|,
\end{equation}
and in 2D, it is
\begin{equation}\label{eq:fnd}
f_{2D}\left(y\right) = 
\begin{cases} 
    \mathbf{E}\left(y\right), &  y \leq 1, \\
    y 
    \left[ 
        \mathbf{E} 
        \left(
            \frac{1}{y} 
        \right)
        -
        \left(
            1 - \frac{1}{y^2}
        \right)
        \mathbf{K}
        \left(
            \frac{1}{y}
        \right)
       \right]  ,& y \geq 1. 
\end{cases}
\end{equation}
The functions $\mathbf{K}(y)$ and $ \mathbf{E}(y) $ are complete elliptic integrals of the first and second kind, respectively. These are implemented in SciPy as \emph{scipy.special.ellipk} and \emph{scipy.special.ellipe}.

\section{Hartree-Fock Stability}
The HF stability conditions (to my knowledge) were presented by Thouless \cite{Thouless1972} ($1^{st}$ Ed. 1960). More general conditions were presented by Adams\cite{Adams1962}. The stability condition is that the eigenvalues ($\lambda$) given by (p. 41 of \cite{Thouless1972})

@@

\begin{equation}
(\epsilon_a-\epsilon_i)C_{ai}^* + \sum\limits_{\text{j}}^{occ}\sum\limits_{\text{b}}^{vir}
	(\braket{ij|ab}-\braket{ij|ba})C_{bj}
	+(\braket{ib|aj}-\braket{ib|ja})C_{bj}^*
	=\lambda C_{ai}^*,
\end{equation}
are nonnegative. This is equivalent to the RPA oscillation frequencies being all real (p. 115 of \cite{Thouless1972}). The RPA is a linearized approximation to Time-Dependent Hartree-Fock theory. These can be written\cite{Dunning1967}\cite{Seeger1977} in the following form:
\begin{equation}\label{eq:1,3H}
{}^{1,3}\bf{H'}=
\begin{bmatrix}
{}^{1,3}\bf{A'} & {}^{1,3}\bf{B'} \\
\left({}^{1,3}\bf{B'}\right)^* & \left({}^{1,3}\bf{A'}\right)^* \\
\end{bmatrix}
\end{equation}
Where the $1,3$ denote singlet and triplet states, respectively. $\bf{A}$, $\bf{B}$ have dimension $N_{occ}\times N_{vir}$ and are defined as follows:
\begin{subequations}
	\begin{align}
	{}^{1}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} + 2\braket{aj|ib}-\braket{aj|bi}\\
	{}^{3}{A'}_{i\rightarrow a, j\rightarrow b} &= (\epsilon_a-\epsilon_i)\delta_{ij}\delta_{ab} - \braket{aj|bi}\\
	{}^{1}{B'}_{i\rightarrow a, j\rightarrow b} &= 2\braket{ab|ij}-\braket{ab|ji}\\
	{}^{3}{B'}_{i\rightarrow a, j\rightarrow b} &= -\braket{ab|ji}
	\end{align}
\end{subequations}
The matrix given in Equation \ref{eq:1,3H} represents the case of complex RHF stability in the space of complex RHF space (${}^{1}\mathbf{H}$ - internal instability) and in the space of complex UHF space (${}^{3}\mathbf{H}$ - triplet instability). Seeger and Pople \cite{Seeger1977} give both more general formulae, and simplifications in special cases. Figure \ref{fig:stability_table} is a reproduction of the stability conditions they developed, and more details can be found in the paper. 
\begin{figure}[h]
   \includegraphics[width=1.0\linewidth]{figures/seeger_stability.pdf}
  	\caption {Hartree-Fock Stability conditions reproduced from Seeger \& Pople \cite{Seeger1977}}
  	\label{fig:stability_table}
\end{figure}

\section{Davidson's Algorithm}
\subsection{Theory}
Oftentimes we're interested only in the lowest or highest few eigenvalues of a large, symmetric matrix. Matrix eigenvalue problem operation counts scale as \url{~}$O(N^3)$ for an $N \times N$ matrix, while the memory requirement scales quadratically. Iterative subspace methods based off the Rayleigh-Ritz method are popular ways of dealing with this issue. The most famous are the Lanczos (hermitian) and Arnoldi (non-hermitian) algorithms, which build as their subspace a Krylov space. A more involved algorithm was proposed by Davidson\cite{Davidson1975} and involves building a subspace with correction vectors to be described momentarily. But first, a few definitions are introduced to define the notation to be used.

@@

\subsection{Implementation}
\begin{algorithm}[H]
\caption{Davidson's Algorithm}\label{davidson}
\begin{algorithmic}[1]
\State create \bf{V}
\For {$i = 0, max\text{ }iterations$}
\State $\bf{\tilde{A}} = \bf{V}^\dag\bf{AV}$
\State $\tilde{\lambda},\mathbf{\tilde{x}} = eigensystem(\bf{\tilde{A}})$
\State sort $\tilde{\lambda},\mathbf{\tilde{x}}$
\If {$i > \textit{stringlen}$} \Return false
\EndIf
\State $j \gets \textit{patlen}$
\If {$\textit{string}(i) = \textit{path}(j)$}
\State $j \gets j-1$.
\State $i \gets i-1$.
\State \textbf{goto} \emph{loop}.
\State \textbf{close};
\EndIf
\State $i \gets i+\max(\textit{delta}_1(\textit{string}(i)),\textit{delta}_2(j))$.
\State \textbf{goto} \emph{top}.
\EndFor
